---
title: 'PSYC 5670: Homework 3'
author: "J DeVaney"
output: 
    html_document: default
---

```{r include = F}
library(haven)
library(here)
library(vtable)
library(dplyr)
library(ggplot2)
library(lme4)
library(lmerTest)
library(nlme)
library(sjPlot)
library(Hmisc)
library(stargazer)
library(knitr)
```

#### Question 1:
```{r echo = T}
#library chunk not included
setwd("~/Desktop/Files/HLM/datasets/")
ext <- read_sas("external_pp.sas7bdat")

#converting ID to factor
ext$ID <- as.factor(ext$ID)
```

#### Question 2:
```{r}
#descriptives table for EXTERNAL
ext %>%
  group_by(TIME) %>%
  summarise(n = n(), mean = mean(EXTERNAL), sd = sd(EXTERNAL), min = min(EXTERNAL), max = max(EXTERNAL)) %>%
  round(digits = 2) %>% 
  kable()
```


#### Question 3:

L1: EXTERNAL~ij~ = B~0j~ + r~ij~

L2: B~0j~ = Gamma~00~ + TIME~j~


Reduced: EXTERNAL~ij~ = Gamma~00~ + TIME~j~ + r~ij~


Gamma~00~: marginal mean of EXTERNAL

TIME~j~: deviation of cluster mean for given timepoint from grand mean

r~ij~: deviation of ith response from jth cluster mean


#### Question 4:
```{r}
ext %>% 
  ggplot(aes(x = TIME, y = EXTERNAL, color = FEMALE)) +
  geom_smooth(method = lm, se =F) +
  facet_wrap(ext$ID)
```

Facet by ID makes obvious the variation in slope and intercept for each person. The arrangement makes it difficult to interpret overall differences between male/female intercepts and slopes. A multilevel model will likely be able to more effectively account for variance, and, as seen in the plots in the next question, will reflect a difference in the intercepts and slopes between males and females.

#### Question 5:
```{r}
p1 <- filter(ext, FEMALE == 1) %>% 
  ggplot(aes(x = TIME, y = EXTERNAL)) +
  geom_smooth(method = lm, se = F, color = "red") +
  ylim(10, 15)
  
p2 <- filter(ext, FEMALE == 0) %>% 
  ggplot(aes(x = TIME, y = EXTERNAL)) +
  geom_smooth(method = lm, se = F, color = "blue") +
  ylim(10, 15)

cowplot::plot_grid(p1, p2)
```

#### Question 6
```{r}
m0 <- lmer(data = ext, EXTERNAL ~ 1 + TIME + (1 + TIME|ID),
           control = lmerControl(optimizer = "bobyqa",
                       optCtrl = list(maxfun = 200000)))
tab_model(m0, show.aic = T, show.r2 = F, show.ci = F, show.se = T)
```



#### Question 7

```{r}
m1 <- lmer(data = ext, EXTERNAL ~ 1 + FEMALE + (1 + TIME|ID),
           control = lmerControl(optimizer = "bobyqa",
                       optCtrl = list(maxfun = 200000)))
```


#### Question 8

```{r}
m2 <- lmer(data = ext, EXTERNAL ~ 1 + TIME*FEMALE + (1 + TIME|ID),
           control = lmerControl(optimizer = "bobyqa",
                       optCtrl = list(maxfun = 200000)))
tab_model(m1, m2, show.aic = T, show.r2 = F, show.ci = F, show.se = T)
```


#### Question 9

```{r}
#ranef needs to be on m1 (random slope for time)
options(scipen = 99999)
mEB <- lmer(data = ext, EXTERNAL ~ 1 + (1|ID), na.action = na.omit)

#from viewing summary for list indices
EB_int <- ranef(mEB)$ID + summary(mEB)[[10]][1]

#index for adding to first column (intercept) and add negative coefficient (slope for time) to time slopes

contrasts(ext$ID) <- contr.sum
```

```{r echo = T, results = 'hide'}
lm(data = ext, EXTERNAL ~ ID) %>% summary()
#lengthy output hidden
```

#### Question 10

```{r}
# p3 <- EBdf %>% 
#   ggplot(aes(x = EBdf[,1], y = EBdf[,2])) +
#   geom_point()
```

#### Question 11

```{r}
hist1 <- ggplot(data = EB_int, aes(x = EB_int[,1])) +
  geom_histogram()
hist1
# hist2 <- ggplot(data = EB_int, aes(x = EB_int[,2])) +
#   geom_histogram()
```

The first histogram of the intercepts shows a positive skew, giving reason to be concerned about violating an assumption of normally distributed random effects.