---
title: "HLM 12-5"
author: "J DeVaney"
output: html_document
---

##### data load
```{r}
library(here)
library(haven)
df <- read_sas(here("datasets", "middleschool.sas7bdat"))
```

*Offset slope approach:*

create another time variable; starts incrementing only once second epoch is entered.

parameter for new time variable estimates difference in rate of change in epoch 2 vs epoch 1


*seperate slope approach:*

build two new time variables

one only increments during epoch 1; stops counting (ex.: 1, 2, 3, 4, 4, 4, 4, 4)

second var is like offset slope approach (ex.: 0, 0, 0, 0, 1, 2, 3, 4)

#### using these ideas
```{r}
library(ggplot2)
library(dplyr)
library(lme4)
library(nlme)

options(scipen = 10)
```

```{r}
with(df, table(grade, middleschool))

#want to create variable to allow intercept change (mean shift) to occur

df$ms <- case_when(
  df$grade %in% 1:5 ~ 0,
  df$grade %in% 6:9 ~ 1
)
```

```{r}
m1 <- lmer(data = df,
           math_theta ~ grade + ms + middleschool +
           (1 + grade + middleschool | id),
          control = lmerControl(optimizer = "bobyqa",
                                optCtrl = list(maxfun=200000)))
summary(m1)
```

midschool slope is 4.15 + -2.32; rate of change is down -2.32

```{r}
df$time1 <- case_when(
  df$grade %in% 1:5 ~ df$grade - 1,
  df$grade %in% 6:9 ~ 4
)

with(df, table(grade, time1))

df$time2 <- case_when(
  df$grade %in% 1:5 ~ 0,
  df$grade %in% 6:9 ~ df$grade - 4
)

with(df, table(grade, time2))
```

```{r}
m2 <- lmer(data = df,
           math_theta ~ time1 + ms + middleschool +
           (1 + time1 + middleschool | id),
          control = lmerControl(optimizer = "bobyqa",
                                optCtrl = list(maxfun=200000)))
summary(m2)
```

middleschool now the actual slope during midschool
significance is difference from slope and zero slope(doesn't tell us much)

*extra stuff:*

all that we have talked about can be applied where outcome variable is not normally distributed